Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-09-09 paper/ms.tex

@@

EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
NASA Astrobiology Institute’s Virtual Planetary Laboratory, supported
by NASA under cooperative agreement NNH05ZDA001C.

>>>>

\section{Solving for log likelihood with extended matrix}

Here we summarize \citet{Ambikasaran2015}, and then point to appendix
for how to do this with oscillating kernels.

\section{Training}

May use derivative of likelihood + BFGS-B to optimize kernel parameters.

Q: How do we determine how many kernels to use?

\section{Prediction/interpolation}

Precition/interpolation involves constructing a matrix of covariances
between the (noisy) training set and the times at which we wish to
predict or interpolate data points.   If our training times are given by
${\bf t}$, while the predicted times are given by ${\bf t}_*$, then
mean values at the predicted times are given by (RW 2.23):
\begin{eqnarray}
\bar f_{*,k} &=& \sum_j K(t_{*,k},t_j) b_j,\\
&=& \sum_j \sum_p \alpha_p e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
&=& \sum_p \alpha_p \sum_j e^{-\beta_p \vert t_{*,k}-t_j\vert} b_j,\\
y_k &=& \sum_j K(t_j,t_k) b_j,
\end{eqnarray}
where the latter equation may be solved for ${\bf b}$ using the extended
matrix formalism.  Let the length of ${\bf t}_*$ be $M$; then, the
matrix $K({\bf t}_*,{\bf t})$ has a size $M \times N$, and so a total
of $MN$ multiplications is required to obtain the predicted values.
If $M$ is large, this can become quickly prohibitive.  It turns out
that the structure of the sum of exponential kernels may be exploited
to obtain the predicted mean in of order $p(M+N)+p^2N$ multiplications,
i.e.\ a linear scaling with the total number of data points, which can 
be much fewer if $M$ is large.

The procedure works as follows:
\begin{itemize}
\item Sort ${\bf t}$ and ${\bf t}_*$ in time order.
\item Starting with the smallest value of ${\bf t}_*$, $t_{*,1}$,
compute:
\begin{eqnarray} \label{prediction}
\bar f_{*,k} &=& \sum_p \alpha_p \left[\epsilon^+_{p,k} + \epsilon^-_{p,k}\right]\\
\epsilon^+_{p,k} &=& \sum_{j \forall t_{*,k} > t_j} e^{-\beta_p (t_{*,k}-t_j)} b_j\\
\epsilon^-_{p,k} &=& \sum_{j \forall t_{*,k} \le t_j} e^{-\beta_p (t_j-t_{*,k})} b_j,
\end{eqnarray}
where we have separated out the training times before and after the predicted time $t_{*,k}$.

\item Next, we update the coefficients recursively:
\begin{eqnarray}
\epsilon^+_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^+_{p,k-1} + \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_{*,k}-t_j)} b_j\right]\\
\epsilon^-_{p,k} &=& e^{-\beta_p (t_{*,k}-t_{*,k-1})} \left[\epsilon^-_{p,k-1} - \sum_{j \forall t_{*,k} > t_j \ge t_{*,k-1}} e^{-\beta_p (t_j-t_{*,k})} b_j\right].
\end{eqnarray}
With these updates, we can then use equation \ref{prediction} to compute the expected
mean at $t_{*,k}$ with an additional $p$ operations.  Since there are $N$ values of
$t_k$, the number of operations needed to update $\epsilon$ is of order $p(N+M)$, and
so this procedure avoids a heavy computational burden.

The variance may be handled in a similar manner, but with $p^2$ coefficients
to recursively update. For example, if $t_{*,k} > t_j \forall j,k$ then:
\begin{eqnarray}
cov(f_{*,k}) &=&  K(t_{*,k},t_{*,k}) - \sum_p \sum_q \alpha_p \alpha_q e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=& K(0) - \sum_p \sum_q \alpha_p \alpha_q \zeta_{p,q,k},\\
\zeta_{p,q,k} &=&   e^{-\beta_p(t_{*,k}-{\bf t}^T)} K({\bf t},{\bf t})^{-1} e^{-\beta_q (t_{*,k}-{\bf t})},\\
&=&  e^{-(\beta_p+\beta_q)(t_{*,k}-t_{*,k-1})} \zeta_{p,q,k-1}.
\end{eqnarray}
The recursive updates to the $\zeta_{p,q,k}$ parameters can be made at each step.
\end{itemize}

Problem:  if we have the case that the datapoints are interspersed, and if the sign changes,
then we need to resolve $K^{-1} e^{-\beta_q \vert t_{*,k}-{\bf t}\vert}$.  However, maybe we can
just solve for the difference in fewer operations? {\bf TBD}

@@

EA acknowledges support from NASA grants NNX13AF20G, NNX13AF62G, and
NASA Astrobiology Institute's Virtual Planetary Laboratory, supported
by NASA under cooperative agreement NNH05ZDA001C.
